---
title: "lab_03"
author: "Derek Willis"
date: "2023-02-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

library(tidyverse)

library(janitor)
```

## Load data and clean up the column names

Load the two datasets we worked with in the pre_lab: the Maryland payments to vendors and Maryland grants and loans and, using janitor, change the columns so that they are lower-cased and any spaces are replaced with underscores.

```{r}
payments <- read_csv("data/State_of_Maryland_Payments_Data__FY2008_to_FY2023.csv")

md_grants_loans <- read_csv("data/State_of_Maryland_Grant_and_Loan_Data__FY2009_to_FY2022.csv")
```

```{r}
cleaned_md_grants_loans <- md_grants_loans %>%
  clean_names()
```


```{r}
cleaned_payments <- payments %>%
  clean_names()
```

## Answer questions

Q1: You want to know how much money state agencies have paid to the Prince George's County School system, but to do that you'll need to find all of the variations of how the school system might be listed. Write code to find rows where the vendor name contains the county's school system, starting with a broad search and then narrowing it down. When you have found as many as possible, save the result to a new dataframe called `pg_schools`. How many records did you find?

A1: I used as many different variations as possible of the name I could think of after looking for different spellings in the data frame and came to a total amount of 35. I have a feeling this is way off as scrolling through the forst 200 or so rows of the database, I feel like I saw at least 15 to 20 that fit the description. Not sure what I am doing wrong but I feel as if there is an easier way to do this then typing in all these different variations and essentially guessing. No matter what your bound to miss a few doing it this way.

```{r}
pg_schools <- cleaned_payments %>%
  filter(vendor_name == "BOARD OF EDUCATION OF PRINCE GEORGES COUNTY MD" | vendor_name == "PRINCE GEORGES COUNTY PUBLIC SCHOOL" | vendor_name == "PRINCE GEORGES COUNTY PUBLIC SCHOOL SYSTEM" | vendor_name == "PRINCE GEORGE COUNTY PUBLIC SCHOOL" | vendor_name == "PRINCE GEORGE COUNTY PUBLIC SCHOOLS" | vendor_name == "PRINCE GEORGES COUNTY PUBLIC SCHOOLS" | vendor_name == "BOARD OF EDUCATION OF PRINCE GEORGE COUNTY MD" | vendor_name == "BOARD OF EDUCATIONS OF PRINCE GEORGES COUNTY MD" | vendor_name == "BOARD OF EDUCATION OF PRINCE GEORGES COUNTY MD") 
  
nrow(pg_schools)
```

Q2: Using that `pg_schools` dataframe, write code to calculate the total amount of payments to the Prince George's County School system for each fiscal year. Which fiscal year had the largest total amount? What do you notice about the years listed?

A2: 2022 had the largest total amount. I find it interesting the top 2 are 2021 and 2022, and then the next 3 are 2008, 2009 and 2010. Why did spending seemingly decrease after 2010 and what changed after 2020?

```{r}
pg_schools <- cleaned_payments %>%
  filter(vendor_name == "BOARD OF EDUCATION OF PRINCE GEORGES COUNTY MD" | vendor_name == "PRINCE GEORGES COUNTY PUBLIC SCHOOL" | vendor_name == "PRINCE GEORGES COUNTY PUBLIC SCHOOL SYSTEM" | vendor_name == "PRINCE GEORGE COUNTY PUBLIC SCHOOL" | vendor_name == "PRINCE GEORGE COUNTY PUBLIC SCHOOLS" | vendor_name == "PRINCE GEORGES COUNTY PUBLIC SCHOOLS" | vendor_name == "BOARD OF EDUCATION OF PRINCE GEORGE COUNTY MD" | vendor_name == "BOARD OF EDUCATIONS OF PRINCE GEORGES COUNTY MD" | vendor_name == "BOARD OF EDUCATION OF PRINCE GEORGES COUNTY MD") 

pg_schools_total_amount <- cleaned_payments %>%
  filter(vendor_name == "BOARD OF EDUCATION OF PRINCE GEORGES COUNTY MD" | vendor_name == "PRINCE GEORGES COUNTY PUBLIC SCHOOL" | vendor_name == "PRINCE GEORGES COUNTY PUBLIC SCHOOL SYSTEM" | vendor_name == "PRINCE GEORGE COUNTY PUBLIC SCHOOL" | vendor_name == "PRINCE GEORGE COUNTY PUBLIC SCHOOLS" | vendor_name == "PRINCE GEORGES COUNTY PUBLIC SCHOOLS" | vendor_name == "BOARD OF EDUCATION OF PRINCE GEORGE COUNTY MD" | vendor_name == "BOARD OF EDUCATIONS OF PRINCE GEORGES COUNTY MD" | vendor_name == "BOARD OF EDUCATION OF PRINCE GEORGES COUNTY MD") %>%
  group_by(fiscal_year) %>%
  summarise(total_amount = sum(amount, na.rm = TRUE)) %>%
  arrange(desc(total_amount))

head(pg_schools_total_amount)
```
Q3. You're interested in examining the vendors paid for capital projects at the University of Maryland. Search the payments data to find those payments, then write code to produce a list of each vendor and the total amount it received. Look carefully at that list - what cleaning will it require to ensure that you have only one row per vendor and zip code? Provide at least three examples of vendor names that are similar and should be combined into one. Describe how you might clean that data so that any nearly identical vendors would look identical - what would you change and how?

A3. One problem that initially jumps out to me is all the "CO.", "INC", "LLC" and "COMPANY" at the end of all the vendor names. There is probably the same company listed other places in the data frame with and without the "LLC" for example. One example of vendor names that should be combined into 1 is "WHITING TURNER CONTRACTING CO" and "WHITING TURNER CONTRACTING COMPANY". Another example is "PLANO-COUDON LLC" and "PLANO-CONDON, LLC". Finally, "VWR INTERNATIONAL LLC" and "VWR INTERNATIONAL INC" should be combined into 1. One simple fix that may work for most cases in this data frame is just looking at the first 8 characters in each vendor_name and matching companies accordingly, or dropping the last 4 or 5 characters from each vendor_name. Although I don't think this will combine vendors who shouldn't be, but I still don't think it will combine every company that should... Surely there is a better way... just what I though of from the zip code problem we worked on in the pre lab Tuesday. 

```{r}
capital_umd_payments <- cleaned_payments %>%
  filter(agency_name == "CAPITAL PROJECTS- UNIVERSITY OF MARYLAND") %>%
  group_by(vendor_name) %>%
  summarise(total_amount = sum(amount, na.rm = TRUE)) %>%
  arrange(desc(total_amount))

capital_umd_payments
```

Q4: You want to understand how much state grant money went to stem cell research projects each fiscal year. Search the data to find grants that fit that description. Once you find which columns contain the information you want, your code should standardize the data needed to find just the grants for stem cell research in a single filter. Then create a total amount of money for each fiscal year. Write a sentence summarizing how stem cell research funding has changed over time.

A4:

```{r}
```
