---
title: "lab_11"
author: "derek willis"
date: "2023-04-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

* Our usual libraries for working with data, plus rvest.

## Load libraries and establish settings

**Task** Create a codeblock and load appropriate packages and settings for this lab.

```{r}
library(rvest)
library(tidyverse)
library(janitor)
library(lubridate)
library(dplyr)
```

Let's get to scraping.

## Questions

**Q1**. Scrape the listing of available Maryland state grants at https://grants.maryland.gov/Pages/StateGrants.aspx into a dataframe. You should have three columns, one of which is a date, so make sure the date column has a date datatype. Then, write code to count the number of grants opportunities offered by each organization listed in your dataframe, showing the organization with the most grant opportunities first. Which state agency has the most?

**A1** Maryland Energy Administration

```{r}
md_grants <- "https://grants.maryland.gov/Pages/StateGrants.aspx"
grant_results <- md_grants %>%
  read_html() %>%
  html_table()
  
grant_results <- grant_results[[1]] %>%
  clean_names() %>%
  rename(title = grant_title) %>%
  mutate(due_date = mdy(due_date))

grants_offered <- grant_results %>%
  count(organization) %>%
  arrange(desc(n)) %>%
  rename(amount_offered = n)
```

**Q2** Next, let's scrape the list of press releases from Maryland's Office of the Public Defender, https://www.opd.state.md.us/press-releases. This isn't a table, so you'll need to use `html_elements()` and your browser's inspector and do some clean up on the results. The result should be a dataframe with two columns that contain the date and title, and the date column should have a date datatype. The challenge here is figuring out how to isolate the releases.

When you finish scraping into a dataframe, write code to find the press releases that have the word "police" in the title. How many are there and when was the most recent one?

**A2** "Police" was in the title of the press releases 9 times. The most recent was 2021-06-21.

```{r}
press_release_url <- "https://www.opd.state.md.us/press-releases"

pr_results <- press_release_url %>%
  read_html()

releases_df <- pr_results %>% html_elements('p.font_8') %>% html_text() %>% as_tibble() %>% separate(value, c('date', 'title'), sep=":") %>% drop_na(title) %>% mutate(date = mdy(date))

police_occurences <- grep("police", releases_df$title, ignore.case = TRUE)
amount_police <- length(police_occurences)
most_recent_police_occurence <- max(releases_df$date[police_occurences])
```

**Q3** Sen. Ben Cardin, D-Maryland, has posted hundreds of press releases at https://www.cardin.senate.gov/?post_type=press-releases. It would be great to have all of them in a dataframe that has the following columns: date, title and url.

To do this, you will need to scrape the page's html and save that to a variable, and _then_ extract the dates, titles and urls into _separate_ dataframes using html_elements(). And remember how we turn a list into a dataframe. The function `html_text()` pulls out the contents of a tag, but for urls we want the HTML attribute. Rvest gives you a way to extract the URL from a link; google to find out what it is.

At the end, you'll have three dataframes that you want to combine into a single dataframe. When we want to combine the rows of identical dataframes, we used `bind_rows()`. If you were combining columns instead of rows, there's a similar function. Find out what it is and use it to put all of the dataframes together into a single one.

When you're done, rename the columns so they make sense, then make sure the date column is an actual date.

Finally, tell me what questions you could ask of this data. Be creative.

**A3** It'd be interesting if we had a dataframe like this for U.S Senator's, or the other U.S Senator for MD. Do we see overlap in titles (topics)? It seems like a lot of the titles include "Cardin", mostly as the first word but sometimes elsewhere. What percentage of titles contain Cardin and is their a correlation between the titles that include the word "Cardin" and the ones that don't to 'good' news vs. not so good news...?

In honestly I think my questions are horrednous. Maybe I am jsut brainfarting but I'm not really seeing a lot of interesting questions to ask of this data. By the action of putting it in a dataframe like this, I can only assume we would be doing so to look for some type of pattern in teh data, othewise there would be no point in putting it in a datframe.

Mayb we would be interested in seeing how frequent releases come out, on what days of the week they come out more... but I don't really find these particularly interesting or creative. Sorry :()

```{r}
cardin_release <- "https://www.cardin.senate.gov/?post_type=press-releases"

cardin_results <- cardin_release %>%
  read_html()

cardin_date <- cardin_results %>% html_elements('h5.customBlog_item__date') %>% html_text() %>% as_tibble()
cardin_title <- cardin_results %>% html_elements("h3.customBlog_item__title") %>% html_text() %>% as_tibble()
cardin_url <- cardin_results %>% html_nodes("a.customBlog_item__readMore") %>% html_attr("href") %>% as_tibble()

cardin_combined <- cbind(cardin_date, cardin_title, cardin_url)

colnames(cardin_combined) <- c("date", "title", "url")

cardin_combined <- cardin_combined %>%
  mutate(date = mdy(date))

  



```
