---
title: "lab_01"
author: "Derek Willis"
date: "2023-01-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Turn off scientific notation
options(scipen=999)
```

## About this lab

To complete this lab, you need to: \* run existing code as directed (look for **Task**). \* modify existing code as directed (look for **Task**). \* write code in empty codeblocks provided to answer questions included (look for **Q**). \* write out the answer in the form of a complete sentence in the space given (look for **A**). When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

Remember to add any new verbs or concepts to your class reference notebook!

## Load libraries and establish settings

**Task** Run the codeblock below to load the Tidyverse. To run the codeblock, hit the little green play button in the upper right corner of the gray area (the codeblock) below.

```{r}
# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
```

## Load Data

We're going to load the UMD courses data.

**Task** Run the codeblock below to load the data, which is contained in an RDS file.

```{r}
umd_courses <- read_rds("lab_01.rds")
```
```{r}
spring2018_umd_courses <- read_rds("lab_01.rds")
```
```{r}
spring2020_umd_courses <- read_rds("lab_01.rds")
```

## Answer questions

**Q1** Which department has the highest mean (average) number of seats for its classes? You need to show the department with the highest average first.
**A1** Engineering, Cooperative Education with a mean number of seats of 314.285714

```{r}
umd_courses %>%
  group_by(department) %>%
  summarise(
    mean_seats = mean(seats),
  ) %>%
  arrange(desc(mean_seats))
```

**Q2** Find out the greatest number of times any individual class has been offered in our dataset. You must show the greatest number of classes first.
**A2** There have been 155 classes offered 21 times, all tied for 1st. 

```{r}
umd_courses %>%
  group_by(id) %>%
  summarise(
    count_classes = n()
  )%>%
  arrange(desc(count_classes))
```

**Q3** Count the number of courses for each instructor and department. Which instructor has taught the most? 
**A3** While doing my code and experimenting to try and find the right answer to this question, I originally just grouped by instructors so I could see which instructor has taught the most classes, before involing the 'department' aspect. When doing so, it showed "Jason Kuo" had taught the most classes with 66. 

After adding to group by 'department', it shows "Keith Yearwood" has taught the msot classes in the Geographical Sciences department with 58. This is showing us which instructor has taught the most in a specific deparmtent. Presumably, Jason Kou has taught the most classes, just in mutliple deparments. 

I've left the code to incldue both the departments and instructors, but I believe the absolute correct answer to the question is Jason Kou, although that is not reflected in the code I have left before.

```{r}
umd_courses %>%
  group_by(instructors, department) %>%
  summarise(
    count_classes = n()
  ) %>%
  arrange(desc(count_classes))
```


**Q4** Given the results for Q3, what would have made it easier to find the answer? You don't have to know _how_ to accomplish this; I want you to describe in words what would make the answer easier to see.
**A4** First, it would be nice to abolish all of the "NA" blocks out of our dataset. I know completley removing them from the dataframe would mess everything else up, but it wold be nice to be able to search for what were looking for and have some line of code that says "disregard or don't show me resuts that have "NA" in them." 

Also, I wish I knew how to show deparments on the chart but not have the computer necessarioly group by it. For example for question 3, I wish I could write some code that would not have any "NA"s, shows Jason Kou as first on the list, and then under "deparments", list all the departments in which he has taught a class. If we could simultaneously for eahc instructor: what deparments have the taught at least one class in, how many classes have they taugh in each individual department, and how many total classes have they taught, would be helpful and in a pefect world I would write a code that does that to answer question 3. 

**Q5** Take one of the questions you listed in the final question of the pre_lab and try to answer it here in code. If you don't know how to fully answer it, write what code you can and describe in words what else you would need to do. If you got an answer, is it interesting?
**A5** Writing this before I start, I am going to attempt to try and figure out a way to show which deparments are upward or downward trending based on number of classes available comparing recent years to older years.

Although I wasn't able to answer the questions, I wrote two code blocks to show the deparments with the most classes in spring of 2018 and spring of 2020. I chose two spring semesters as I thought that would be the best way to compare, as using a winter and a spring semester for comparison would not be correct as there is clerly more classes offered in the spring vs winter.

The problem I ran into is I want to be able to take the classes from spring of 2020 for each deparment and subtracting it from the number of classes offered from that department in spring of 2018. For example, Business and Management has actually seen a drop in classes in 2020 compared to 2018. The equation would be 154 - 168 = -14, indicating the Business and Managment department offered 14 less classes in 2020 spring then in 2018 spring.

Departments that had the highest number would be shwoing an upward trend, whereas departments with low/negative numbers would show a trend. I tried to solve this by creating multiple dataset ids, but I am still unsure how to take the two and subtract them into another dataset. 

```{r}
spring2020_umd_courses %>%
  filter(term == 202001) %>%
  group_by(department) %>%
  summarise(
    count_classes = n()
  ) %>%
  arrange(desc(count_classes))
```

```{r}
spring2018_umd_courses %>%
  filter(term == 201801) %>%
  group_by(department) %>%
  summarise(
    count_classes = n()
  ) %>%
  arrange(desc(count_classes))
```


